Identifying supermassive black hole recoil in elliptical galaxies

### 2024-10-21 papers/paper-corekick/paper-corekick.tex
Paper version as submitted to arXiv

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[A. Rawlings et al.]{Alexander Rawlings,$^{1}$\thanks{E-mail: alexander.rawlings@helsinki.fi}
Atte Keitaanranta,$^{1}$
Max Mattero,$^{1}$
Sonja Soininen,$^{1}$
Ruby J. Wright,$^{2,1}$
\newauthor
Noa Kallioinen,$^{3}$
Shihong Liao,$^{4}$
Antti Rantala,$^{5}$
Peter H. Johansson,$^{1}$
Thorsten Naab,$^{5}$ and
\newauthor Dimitrios Irodotou$^{6}$
\vspace*{0.1cm}\\
% List of institutions
$^{1}$
Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland
\\
$^{2}$
International Centre for Radio Astronomy Research (ICRAR), University of Western Australia, Crawley, WA 6009, Australia
\\
$^{3}$
Department of Computer Science, Aalto University, Espoo, Finland
\\
$^{4}$
Key Laboratory for Computational Astrophysics, National Astronomical Observatories, Chinese Academy of Sciences, Beijing 100101, China
\\
$^{5}$
Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1, D-85748 Garching, Germany
\\
$^{6}$
The Institute of Cancer Research, London SW7 3RP, UK
\\
}

@@

% Abstract of the paper
\begin{abstract}
\alex{We have 250 words to use here. Currently 247 excluding the 3 math inlines.}

We study the contribution to stellar core growth in simulations of merging massive ($M_\star>10^9\,\Msun$) elliptical galaxies by a coalesced supermassive black hole (SMBH) displaced by gravitational wave (GW) induced recoil velocity.
With controlled, dense sampling of the SMBH recoil velocity, we find the stellar core originally formed by SMBH binary scouring can grow by a factor of 2-3 with increasing recoil velocity, and a similar growth in the central mass deficit.
Using Bayesian inference we predict the distribution of stellar core sizes formed through this process to peak at $\sim1\,\mathrm{kpc}$.
An orbital decomposition of stellar particles within the core reveals that when the recoil velocity exceeds the velocity dispersion of the core, radial orbits dominate over tube orbits, providing a mass 10\% of the coalesced SMBH mass, whereas tube orbits dominate for the lowest recoil kicks.
A change in orbital structure is reflected in the anisotropy parameter, with a tangential bias present only for kick velocities less than the local stellar velocity dispersion.
With mock integral field unit observations of the stellar line of sight velocity distribution emulating the capabilities of current detectors, we observe a distinct signature in the Gauss-Hermite symmetric deviation coefficient $h_4$ that allows for an estimate of the core size due to binary scouring, irrespective of later evolution due to SMBH recoil.
Our results provide a method to estimate the SMBH recoil magnitude from observations of local elliptical galaxies, and implies these galaxies primarily experienced recoil velocities less than the stellar velocity dispersion of the core.
\end{abstract}

@@

% INTRODUCTION
\section{Introduction}
%There is much evidence for the presence of supermassive black holes (SMBHs) at the centres of massive galaxies, with the shadow of one such SMBH being directly imaged \citep{eht2019}.
%As two galaxies dynamically interact during a galaxy merger, so too is it expected that the residing SMBHs will dynamically interact.
Observations have revealed a clear bimodality in the properties of massive elliptical galaxies, with brighter ellipticals $(M_{B}\lesssim -20.5)$ typically having `cored' flat central surface brightness profiles, boxy isophotes and showing little or no rotation (slow rotators). Fainter intermediate-mass ellipticals $(-20.5 \lesssim M_{B} \lesssim -18.5)$ on the other hand typically show steeper `cuspy' central brightness profiles, more disc-like isophotes and evidence for more rotational support (fast rotators) (e.g \citealt{Kormendy1996,Faber1997,emsellem2007,Kormendy2009,Thomas2014,dullo2014,dullo2019,Quenneville2024}).

@@

In the case of two massive elliptical galaxies undergoing a dry merger, the leading mechanism for stellar core formation is through the interaction of binary supermassive black holes (SMBHs) in the galaxy merger remnant. The merging of SMBHs 
is generally understood in the context of a three stage process \citep{begelman1980}. First, the dynamical friction from the mass overdensity in the wakes of the SMBHs provides a restoring force which brings the SMBHs to the centre of the two interacting galaxies \citep{chandrasekhar1943}, generally torquing the SMBHs to highly radial orbits in the process. Second, after the mass enclosed within the orbit of the two SMBHs is comparable to the sum of the SMBH masses, the SMBHs form a bound binary \citep{merritt2005}. Iterative interactions of the SMBHs with surrounding stars on low angular momentum orbits ejects these stars with high velocity \citep[e.g.][]{hills1980,hills1983,quinlan1996}, thus removing orbital energy and angular momentum from the SMBH binary orbit.
The removal of stars from the central regions by this slingshot mechanism produces a `core' \citep[e.g.][]{lauer1983,lauer1985,kormendy1984}, or depletion, in the luminosity profile of the remnant galaxy \citep[e.g.][]{begelman1980,hills1983,quinlan1996,rantala2018,rantala2024,nasim2021b,khonji2024,partmann2024}, that can in extreme cases extend up to $3\,\mathrm{kpc}$ \citep{postman2012}.
Finally, at very small, sub-parsec separations the SMBH binary loses its remaining orbital energy and angular momentum through gravitational wave (GW) emission, driving the SMBH binary to coalescence \citep{peters1963,peters1964}. The GW emission from such SMBH coalescence events are prime observational targets for ongoing pulsar timing array (PTA) detectors for SMBHs with masses of $M_\bullet \gtrsim 10^9\,\Msun$ \citep{agazie2023, 2023Antoniadis, 2023Xu, zic2023}, and also the forthcoming Laser Interferometer Space Antenna (LISA) mission for SMBHs with slightly lower masses of $M_\bullet \lesssim 10^8\,\Msun$ \citep{amaro-seoane2023}.

The final GW emission from the coalescing SMBH binary occurs anisotropically, and carries with it linear momentum from the system \citep[e.g.][]{gonzalez2007}.
This results in the remnant SMBH recoiling with some velocity directed opposite to the linear momentum of the GW emission \citep{bekenstein1973}, and is termed the kick, or recoil, velocity\footnote{We use the terms `kick' and `recoil' interchangeably in this work.}. 
The kick velocity is dependent on the mass ratio between the SMBHs prior to coalescence, and the magnitude and direction of the angular momentum (spin) vector of each SMBH. 
As reported by \citet{campanelli2007}, symmetries in the masses or spins of the SMBHs suppresses the resulting recoil kick imparted to the coalesced SMBH.
In particular, it is found that asymmetry in the spins of the SMBHs has a larger impact on the magnitude of the recoil kick than asymmetry in the masses \citep{campanelli2007}, with spins which are anti-aligned in the orbital plane and maximal producing the largest recoil kick velocities \citep{gonzalez2007b,tichy2007}.
Numerical relativity studies indicate that while the majority of recoil kicks are of the order of a few hundred kilometres per second, in special configurations the SMBH may be imparted a kick velocity in excess of $2000\,\kmps$, even up to $4000\,\kmps$ \citep{campanelli2007,gonzalez2007b,tichy2007}.
Taking the central escape velocity of a typical massive elliptical galaxy as $\sim2000\,\kmps$, there may be a non-negligible fraction of elliptical galaxies with an SMBH that has escaped the galaxy (e.g. \citealt{madau2004,mannerkoski2022}), though not so many so as to introduce considerable scatter into the observed relation between SMBH mass and stellar velocity dispersion \citep{volonteri2007}.
This is in clear tension with all current observations indicating that no massive elliptical galaxy in the local Universe is without an SMBH, and demands further investigation.

@@

Numerically, the effect of recoiling SMBHs on the formation of cores in massive elliptical galaxies has been 
primarily studied using collisionless simulations. Early work by \citet{boylan2004}
found that a recoiling SMBH induced a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion. In another early study 
\citet{merritt2004} also found that recoiling SMBHs facilitated the formation of cores
and found that ejected SMBHs settled roughly in one orbital period in spherical potentials, with this `infall time' increasing when more realistic triaxial potentials were considered. 
In \citet{gualandris2008} the motion of a settling SMBH in a stellar potential representative of a galaxy merger remnant was found to occur in three stages: first, the SMBH oscillates with decreasing amplitude due to energy dissipation by dynamical friction.
Next, both the SMBH and stellar core oscillate about their common centre of mass when the radial excursion of the SMBH is less than the core radius of the stellar system.
Finally, the SMBH settles into thermal equilibrium with the surrounding stars.
During these three stages, the stellar density core is enhanced by the motion of the SMBH, producing a mass deficit in some cases up to five times the SMBH mass (see also \citealt{merritt2004,boylan2004,partmann2024}).
However, a more recent study by \citet{nasim2021b} on the contribution to stellar core expansion through GW recoil of an SMBH found intriguingly that the greatest contribution to the stellar mass deficit is due to the first excursion of the SMBH following its ejection, with subsequent oscillations only providing a minor contribution to the growth of the stellar core. 
%This result is in tension with the work of \citet{gualandris2008}.

%The effect of recoiling SMBHs have been studied previously, with some of the first explorations presented in \citet{boylan2004} and \citet{merritt2004}.
%In the work of \citet{boylan2004}, a recoiling SMBH was found to induce a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion.
%A recoiling SMBH was also found to induce a stellar density core in the work of \citet{merritt2004}.
%Additionally, the time for a recoiling SMBH to settle after being ejected in a spherical potential was found to be roughly one orbital period when the recoil velocity was less than the velocity dispersion of the bulge.
%This, which are more representative of a galaxy merger remnant than the spherical case.

%More recently, the contribution of stellar core expansion through GW recoil of an SMBH has been considered by \citet{nasim2021b}.
%Intriguingly, the authors find that greatest contribution to the stellar mass deficit is due to the first excursion of the SMBH following its ejection, with subsequent oscillations only providing a minor contribution to the growth of the stellar core.
%This result is in tension with the work of \citet{gualandris2008}.

%The effect of recoiling SMBHs have been studied previously, with some of the first explorations presented in \citet{boylan2004} and \citet{merritt2004}.
%In the work of \citet{boylan2004}, a recoiling SMBH was found to induce a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion.
%A recoiling SMBH was also found to induce a stellar density core in the work of \citet{merritt2004}.
%Additionally, the time for a recoiling SMBH to settle after being ejected in a spherical potential was found to be roughly one orbital period when the recoil velocity was less than the velocity dispersion of the bulge.
%This `infall time' was found to increase when considering triaxial potentials, which are more representative of a galaxy merger remnant than the spherical case.

%In \citet{gualandris2008}, a stellar potential representative of a galaxy merger remnant with SMBH binary core scouring is used to investigate the motion of an SMBH ejected through GW recoil.
%The motion of a settling SMBH was found to occur in three stages: first, the SMBH oscillates with decreasing amplitude due to energy dissipation by dynamical friction.
%Next, both the SMBH and stellar core oscillate about their common centre of mass when the radial excursion of the SMBH is less than the core radius of the stellar system.
%Finally, the SMBH settles into thermal equilibrium with the surrounding stars.
%During these three stages, the stellar density core is enhanced by the motion of the SMBH (in agreement with \citet{merritt2004} and \citet{boylan2004}), with the mass deficit in some cases reaching up to five times the SMBH mass being produced.

%Work by \citet{blecha2008} considered the effect of recoiling SMBHs in an analytical model of an isolated galaxy comprised of a multicomponent system of a stellar bulge and gaseous disc.
%An interesting result was the mass growth of the ejected SMBH as it settled at the centre of the potential, with the mass growth accumulating to some 10\% of the initial SMBH mass, irrespective of the recoil velocity.
%Recoiling SMBHs in gas-rich systems was further explored in \citet{blecha2011} using hydrodynamic simulations of isolated galaxy mergers.
%In this work, seed SMBHs were allowed to grow through gas accretion prior to merging, with the majority of final SMBH masses at the time of merger sampling the $10^7\,\Msun \lesssim M_\bullet \lesssim 10^8\,\Msun$ mass regime.

%The impact of recoiling SMBHs on the growth and dynamics of SMBHs in hydrodynamical simulations have also been studied using both isolated galaxy models consisting of a stellar bulge and a gaseous disc \citep{blecha2008}, as well as in isolated gas-rich galaxy mergers \citep{blecha2011}. %The study of the isolated galaxy model found that 
%the mass growth of the ejected SMBH as it settled at the centre of the potential, with the mass growth accumulating to some 10\% of the initial SMBH mass, irrespective of the recoil velocity. 
%In the merger simulations, the seed SMBHs were allowed to grow through gas accretion prior to merging, with the majority of the final SMBH masses at the time of merger sampling the $10^7\,\Msun \lesssim M_\bullet \lesssim 10^8\,\Msun$ mass regime. The simulations also showed that recoil events typically reduce the lifetimes of bright active galactic nuclei (AGNs), with most kicked AGNs confined to the central few kpc of the galaxy. Whilst the results of \citet{blecha2011} provide valuable insight into the occurrence of variable AGN variability due to an oscillating SMBH plunging through a gaseous disc, we stress that the results are applicable to a different mass regime of galaxies and SMBHs than what we focus on in this work. In addition, the formation of stellar cores which is primarily driven by complex three-body interactions between the SMBHs and the stellar component remains largely unresolved in simulations that include gravitational softening.

@@

%In this study, we wish to consider SMBH recoil by GW emission in massive elliptical galaxies.
%The aim of this work is twofold.
%We first aim to determine how the velocity with which the SMBH is kicked affects the mass distribution of the surrounding stellar environment, and the likelihood of the magnitude of these effects.
%We then aim to determine if a recoiling SMBH imparts a signature on the stellar kinematics of the remnant galaxy, and if such a signature could be observed with current detectors.

This paper is divided as follows: we present the numerical simulations used in our investigation in \autoref{sec:num_sims}.
We discuss the motion of the recoiling SMBHs in \autoref{sec:motion}, and the resulting stellar density profiles of the remnant galaxy in \autoref{sec:density}.
We investigate the distribution of stellar orbits as a function of the kick velocity in \autoref{sec:orbits} and construct mock stellar kinematic observations to be compared with observational data in \autoref{sec:kinematics}.
Finally, we discuss of our results in \autoref{sec:discussion} and present 
our conclusions in \autoref{sec:conclusions}, respectively.

@@

\subsection{Simulation Code}
To investigate the effect of the recoiling SMBH on the host galaxy, and potential observational signatures of this, we run a number of numerical simulations of an idealised galaxy merger setting using the new public version of the \ketju{}\footnote{\url{https://www.mv.helsinki.fi/home/phjohans/group-website/research/ketju/}} code \citep{rantala2017,rantala2020,mannerkoski2023} coupled with \gadget{} \citep{springel2021}.
\ketju{} integrates the dynamics of SMBHs, and stellar particles in a small region thrice the stellar softening length around them (i.e. the \ketju{} region), to high accuracy using the algorithmically regularised integrator \textsc{mstar} \citep{rantala2020}.
The dynamics of stellar particles beyond this small region of high integration accuracy, and all DM particles, is computed with the \gadget{} fast multiple method (FMM) with second order multipoles. 
Additionally, we use hierarchical time integration, which allows for mutually symmetric interactions and manifest momentum conservation. 
In this study, we do not use comoving integration, as we are investigating an idealised system.
\ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014} and the fitting formula of \citet{zlochower2015} for recoil kick velocity, making \ketju{} a particularly well-suited code for investigating the consequence of SMBH recoil self-consistently in a galaxy merger environment.

@@

As we are interested in the evolution of the galaxy merger \textit{remnant} following SMBH coalescence, the initial conditions are constructed such that at the time of SMBH coalescence (\autoref{ssec:mergers}) the galaxy remnant agrees with local scaling relations.
Specifically, we ensure that our remnant agrees with the half-light -- stellar mass data presented in \citet{sahu2020}, and lies within the $1\sigma$ prediction interval of the SMBH mass -- stellar velocity dispersion data given in \citet{vandenbosch2016}.

\subsection{Merger Simulations}\label{ssec:mergers}
We merge two independently Monte Carlo sampled galaxy ICs, creating an equal mass galaxy merger with a near-radial orbit. 
The merger orbit has an initial separation consistent with \citet{rantala2018} of $D=30\,\mathrm{kpc}$, however the initial eccentricity is set to $e_0=0.97$, and the first pericentre distance adjusted accordingly to $r_\mathrm{peri}=D(1-e_0^2) \simeq 2\,\mathrm{kpc}$ \citep{khochfar2006,rantala2017}.
The merger configuration results in a rapid coalescence of the SMBH binary, however we note that the merger timescale we observe is driven by stochasticity in the binary eccentricity \citep{nasim2020,rawlings2023}.
The resulting coalesced SMBH has a mass $M_\bullet = 2M_{\bullet,0} = 5.86\times10^9\,\Msun$.
We then select the snapshot just prior ($\sim8\,\mathrm{Myr}$) to the GW-driven SMBH merger, and generate 31 `child' simulations, where each child has a unique gravitational recoil kick velocity $\vk$ prescribed along the $x$-axis\footnote{As the $x$-axis is in the global coordinate frame, the direction of the kick is essentially random with respect to the angular momentum and inertia tensors. To test the effect of differing kick directions on the merger remnant, we test directing the recoil kick along the global $y$-axis, and find very similar evolution to the $x$-axis case.}, ranging from $0\,\kmps$ to $1800\,\kmps$ (inclusive), in $60\,\kmps$ increments.
%At the moment of the merger, the centre of mass (CoM) velocity of the SMBH binary is $v_{\mathrm{bin},\mathrm{CoM}}\sim[4.2, 1.1, 1.6]\,\kmps$ in the global coordinate frame.
The upper limit on $\vk$ is chosen to match the escape velocity of the centre of the merger remnant, $v_\mathrm{esc}$.
We additionally run one simulation with a recoil kick above $v_\mathrm{esc}$, with $\vk=2000\,\kmps$, to test the effect of rapidly removing an SMBH.

For all simulations, snapshots are produced every $5\,\mathrm{Myr}$. Interactions between stellar particles are softened with a softening length of $\varepsilon = 2.5\,\mathrm{pc}$, DM with a softening length of $\varepsilon = 300\,\mathrm{pc}$, and the \ketju{} region radius is set to $r_{\rm ketju}=3\varepsilon = 7.5\,\mathrm{pc}$.
Within the \ketju{} region, dynamical interactions between stellar particles and SMBHs are unsoftened.
For the \gadget{} component of the simulation, the integration error tolerance is set to 0.002, whereas the Gragg-Bulirsch-Stoer (GBS) tolerance for \ketju{} is set to $10^{-8}$.

@@

The settling times are shown in the lower panel of \autoref{fig:infall}. For recoil velocities $\vk \leq 240\,\kmps$ the kick is not enough to displace the SMBH beyond the binary-scoured core and the SMBH is considered settled in less than $100\,\mathrm{Myr}$ after SMBH binary coalescence. 
As the kick velocity is increased further, the settling time begins to increase exponentially. 
The slight deviations from an exponential increase are caused by numerical noise, which prevents the median velocity being below $V_\mathrm{RMS}$.

The SMBHs which received a kick velocity larger than $1020\,\kmps$ did not settle back to the centre of the galaxy within the maximum simulation time of $3t_\mathrm{dyn} = 3\,\mathrm{Gyr}$, and already reach maximum displacements in excess of $>60\,\mathrm{kpc}$ during their first excursion from the centre. 
We expect that any SMBH with such a large displacement from its host galaxy would not be observationally identified with that galaxy, and thus we do not include simulations with $\vk>1020\,\kmps$ in this analysis. The used limit for maximum displacement is a few times larger than the largest suggested offset of $19.4\,\mathrm{kpc}$ from observations \citep{barrows2016}.
We do however analyse the $\vk=2000\,\kmps$ simulation, which is above the escape velocity of the galaxy, as a proxy for an SMBH that is instantaneously removed from the simulation.
Importantly, an escaping SMBH never completes a single oscillation about the centre of the galaxy.

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[A. Rawlings et al.]{Alexander Rawlings,$^{1}$\thanks{E-mail: alexander.rawlings@helsinki.fi}
Atte Keitaanranta,$^{1}$
Max Mattero,$^{1}$
Sonja Soininen,$^{1}$
Ruby J. Wright,$^{2,1}$
\newauthor
Noa Kallioinen,$^{3}$
Shihong Liao,$^{4}$
Antti Rantala,$^{5}$
Peter H. Johansson,$^{1}$
Thorsten Naab,$^{5}$ and
\newauthor Dimitrios Irodotou$^{6}$
\vspace*{0.1cm}\\
% List of institutions
$^{1}$
Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland
\\
$^{2}$
International Centre for Radio Astronomy Research (ICRAR), University of Western Australia, Crawley, WA 6009, Australia
\\
$^{3}$
Department of Computer Science, Konemiehentie 2, 02150, Aalto University, Espoo, Finland
\\
$^{4}$
Key Laboratory for Computational Astrophysics, National Astronomical Observatories, Chinese Academy of Sciences, Beijing 100101, China
\\
$^{5}$
Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1, D-85748 Garching, Germany
\\
$^{6}$
The Institute of Cancer Research, 123 Old Brompton Road, London SW7 3RP, United Kingdom
\\
}

@@

% Abstract of the paper
\begin{abstract}
We study stellar core growth in simulations of merging massive ($M_\star>10^{11}\,\Msun$) elliptical galaxies by a supermassive black hole (SMBH) displaced by gravitational wave induced recoil velocity.
With controlled, dense sampling of the SMBH recoil velocity, we find the core radius originally formed by SMBH binary scouring can grow by a factor of 2-3 when the recoil velocity exceeds $\sim50$ per cent of the central escape velocity, and the mass deficit grows by up to a factor of $\sim4$.
Using Bayesian inference we predict the distribution of stellar core sizes formed through this process to peak at $\sim1\,\mathrm{kpc}$.
An orbital decomposition of stellar particles within the core reveals that radial orbits dominate over tube orbits when the recoil velocity exceeds the velocity dispersion of the core, whereas tube orbits dominate for the lowest recoil kicks.
A change in orbital structure is reflected in the anisotropy parameter, with a central tangential bias present only for recoil velocities less than the local stellar velocity dispersion.
Emulating current integral field unit observations of the stellar line-of-sight velocity distribution, we uncover a distinct signature in the Gauss-Hermite symmetric deviation coefficient $h_4$ that uniquely constrains the core size due to binary scouring.
This signature is insensitive to the later evolution of the stellar mass distribution due to SMBH recoil.
Our results provide a novel method to estimate the SMBH recoil magnitude from observations of local elliptical galaxies, and implies these galaxies primarily experienced recoil velocities less than the stellar velocity dispersion of the core.
\end{abstract}

@@

% INTRODUCTION
\section{Introduction}
Observations have revealed a clear bimodality in the properties of massive elliptical galaxies, with brighter ellipticals $(M_{B}\lesssim -20.5)$ typically having `cored' flat central surface brightness profiles, boxy isophotes and showing little or no rotation (slow rotators). Fainter intermediate-mass ellipticals $(-20.5 \lesssim M_{B} \lesssim -18.5)$ on the other hand typically show steeper `cuspy' central brightness profiles, more disc-like isophotes and evidence for more rotational support (fast rotators) (e.g. \citealt{Kormendy1996,Faber1997,emsellem2007,Kormendy2009,Thomas2014,dullo2014,dullo2019,Quenneville2024}).

@@

In the case of two massive elliptical galaxies undergoing a dry merger, the leading mechanism for stellar core formation is through the interaction of binary supermassive black holes (SMBHs) in the galaxy merger remnant. The merging of SMBHs 
is generally understood in the context of a three stage process \citep{begelman1980}. First, the dynamical friction from the mass overdensity in the wakes of the SMBHs provides a restoring force which brings the SMBHs to the centre of the two interacting galaxies \citep{chandrasekhar1943}, generally torquing the SMBHs to highly radial orbits in the process. Second, after the mass enclosed within the orbit of the two SMBHs is comparable to the sum of the SMBH masses, the SMBHs form a bound binary \citep{merritt2005}. Iterative interactions of the SMBHs with surrounding stars on low angular momentum orbits eject these stars with high velocity \citep[e.g.][]{hills1980,hills1983,quinlan1996}, thus removing orbital energy and angular momentum from the SMBH binary orbit.
The removal of stars from the central regions by this slingshot mechanism produces a `core' \citep[e.g.][]{lauer1983,lauer1985,kormendy1984}, or depletion, in the luminosity profile of the remnant galaxy \citep[e.g.][]{begelman1980,hills1983,quinlan1996,rantala2018,rantala2024,nasim2021b,khonji2024,partmann2024}, that can in extreme cases extend up to $3\,\mathrm{kpc}$ \citep{postman2012}.
Finally, at very small, sub-parsec separations the SMBH binary loses its remaining orbital energy and angular momentum through gravitational wave (GW) emission, driving the SMBH binary to coalescence \citep{peters1963,peters1964}. The GW emission from such SMBH coalescence events are prime observational targets for ongoing pulsar timing array (PTA) detectors for SMBHs with masses of $M_\bullet \gtrsim 10^9\,\Msun$ \citep{agazie2023, antoniadis2024, 2023Xu, zic2023}, and also the forthcoming Laser Interferometer Space Antenna (LISA) mission for SMBHs with slightly lower masses of $M_\bullet \lesssim 10^8\,\Msun$ \citep{amaro-seoane2023}.

The final GW emission from the coalescing SMBH binary occurs anisotropically, and carries with it linear momentum from the system \citep[e.g.][]{gonzalez2007}.
This results in the remnant SMBH recoiling with some velocity directed opposite to the linear momentum of the GW emission \citep{bekenstein1973}, and is termed the kick, or recoil, velocity\footnote{We use the terms `kick' and `recoil' interchangeably in this work.}. 
The kick velocity is dependent on the mass ratio between the SMBHs prior to coalescence, and the magnitude and direction of the angular momentum (spin) vector of each SMBH. 
As reported by \citet{campanelli2007}, symmetries in the masses or spins of the SMBHs suppresses the resulting recoil kick imparted to the coalesced SMBH.
In particular, it is found that asymmetry in the spins of the SMBHs has a larger impact on the magnitude of the recoil kick than asymmetry in the masses \citep{campanelli2007}, with spins which are anti-aligned in the orbital plane and maximal producing the largest recoil kick velocities \citep{gonzalez2007b,tichy2007}.
Numerical relativity studies indicate that while the majority of recoil kicks are of the order of a few hundred kilometres per second, in special configurations the SMBH may be imparted a kick velocity in excess of $2000\,\kmps$, even up to $4000\,\kmps$ \citep{campanelli2007,gonzalez2007b,tichy2007}.
Taking the central escape velocity of a typical massive elliptical galaxy as $\sim2000\,\kmps$, there may be a non-negligible fraction of elliptical galaxies with an SMBH that has escaped the galaxy (e.g. \citealt{madau2004,mannerkoski2022}), though not so many so as to introduce considerable scatter into the observed relation between SMBH mass and stellar velocity dispersion \citep{volonteri2007}.

@@

Numerically, the effect of recoiling SMBHs on the formation of cores in massive elliptical galaxies has been 
primarily studied using collisionless simulations. Early work by \citet{boylan2004}
found that a recoiling SMBH induced a core in the stellar density profile, particularly when the SMBH remained bound to the bulge of the host galaxy and had a recoil velocity comparable to the bulge velocity dispersion. In another early study 
\citet{merritt2004} also found that recoiling SMBHs facilitated the formation of cores
and found that ejected SMBHs settled roughly in one orbital period in spherical potentials, with this `infall time' increasing when more realistic triaxial potentials were considered. 
In \citet{gualandris2008} the motion of a settling SMBH in a stellar potential representative of a galaxy merger remnant was found to occur in three stages: first, the SMBH oscillates with decreasing amplitude due to energy dissipation by dynamical friction.
Next, both the SMBH and stellar core oscillate about their common centre of mass when the radial excursion of the SMBH is less than the core radius of the stellar system.
Finally, the SMBH settles into thermal equilibrium with the surrounding stars.
During these three stages, the stellar density core is enhanced by the motion of the SMBH, producing a mass deficit in some cases up to five times the SMBH mass (see also \citealt{merritt2004,boylan2004,partmann2024}).
However, a recent study by \citet{nasim2021b} on the contribution to stellar core expansion through GW recoil of an SMBH found intriguingly that the greatest contribution to the stellar mass deficit is due to the first excursion of the SMBH following its ejection, with subsequent oscillations only providing a minor contribution to the growth of the stellar core.

@@

This paper is divided as follows: we present the numerical simulations used in our investigation in \autoref{sec:num_sims}.
We discuss the motion of the recoiling SMBHs in \autoref{sec:motion}, and the resulting stellar density profiles of the remnant galaxy in \autoref{sec:density}.
We investigate the distribution of stellar orbits as a function of the kick velocity in \autoref{sec:orbits} and construct mock stellar kinematic observations to be compared with observational data in \autoref{sec:kinematics}.
Finally, we discuss our results in \autoref{sec:discussion} and present 
our conclusions in \autoref{sec:conclusions}, respectively.

@@

\subsection{Simulation Code}
To investigate the effect of the recoiling SMBH on the host galaxy, and potential observational signatures of this, we run a number of numerical simulations of an idealised galaxy merger setting using the new public version of the \ketju{}\footnote{\url{https://www.mv.helsinki.fi/home/phjohans/group-website/research/ketju/}} code \citep{rantala2017,rantala2020,mannerkoski2023} coupled with \gadget{} \citep{springel2021}.
\ketju{} integrates the dynamics of SMBHs, and stellar particles in a small region three times the stellar softening length around them (i.e. the \ketju{} region), to high accuracy using the algorithmically regularised integrator \textsc{mstar} \citep{rantala2020}.
The dynamics of stellar particles beyond this small region of high integration accuracy, and all DM particles, is computed with the \gadget{} fast multiple method (FMM) with second order multipoles. 
Additionally, we use hierarchical time integration, which allows for mutually symmetric interactions and manifest momentum conservation. 
In this study, we do not use comoving integration, as we are investigating an idealised system.
\ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014} and the fitting formula of \citet{zlochower2015} for recoil kick velocity, making \ketju{} a particularly well-suited code for investigating the consequence of SMBH recoil self-consistently in a galaxy merger environment.

@@

As we are interested in the evolution of the galaxy merger \textit{remnant} following SMBH coalescence, the initial conditions are constructed such that at the time of SMBH coalescence (\autoref{ssec:mergers}) the galaxy remnant agrees with local scaling relations.
Specifically, we ensure that our remnant agrees with the half-light -- stellar mass data presented in \citet{sahu2020}, and lies within the $1\sigma$ prediction interval of the SMBH mass -- stellar velocity dispersion data as given in \citet{vandenbosch2016}.

\subsection{Merger Simulations}\label{ssec:mergers}
We merge two independently Monte Carlo sampled galaxy ICs, creating an equal-mass galaxy merger with a near-radial orbit. 
The merger orbit has an initial separation consistent with \citet{rantala2018} of $D=30\,\mathrm{kpc}$, however the initial eccentricity is set to $e_0=0.97$, and the first pericentre distance adjusted accordingly to $r_\mathrm{peri}=D(1-e_0^2) \simeq 2\,\mathrm{kpc}$ \citep{khochfar2006,rantala2017}.
The merger configuration results in a rapid coalescence of the SMBH binary, however we note that the merger timescale we observe is driven by stochasticity in the binary eccentricity \citep{nasim2020,rawlings2023}.
The resulting coalesced SMBH has a mass $M_\bullet = 2M_{\bullet,0} = 5.86\times10^9\,\Msun$.
We then select the snapshot just prior ($\sim8\,\mathrm{Myr}$) to the GW-driven SMBH merger, and generate 31 `child' simulations, where each child has a unique gravitational recoil kick velocity $\vk$ prescribed along the $x$-axis\footnote{As the $x$-axis is in the global coordinate frame, the direction of the kick is essentially random with respect to the angular momentum and inertia tensors. To test the effect of differing kick directions on the merger remnant, we test directing the recoil kick along the global $y$-axis, and find very similar evolution to the $x$-axis case.}, ranging from $0\,\kmps$ to $1800\,\kmps$ (inclusive), in $60\,\kmps$ increments.
The upper limit on $\vk$ is chosen to match the escape velocity of the centre of the merger remnant, $v_\mathrm{esc}=1800\,\kmps$.
We additionally run one simulation with a recoil kick above $v_\mathrm{esc}$, with $\vk=2000\,\kmps$, to test the effect of rapidly removing an SMBH.

For all simulations, snapshots are produced every $5\,\mathrm{Myr}$. Interactions between stellar particles are softened with a softening length of $\varepsilon_\star = 2.5\,\mathrm{pc}$, DM is softened with a softening length of $\varepsilon_\mathrm{DM} = 300\,\mathrm{pc}$, and the \ketju{} region radius is set to $r_{\rm ketju}=3\varepsilon_\star = 7.5\,\mathrm{pc}$.
Within the \ketju{} region, dynamical interactions between stellar particles and SMBHs, and SMBHs-SMBHs are unsoftened.
For the \gadget{} component of the simulation, the integration error tolerance is set to 0.002, whereas the Gragg-Bulirsch-Stoer (GBS) tolerance for \ketju{} is set to $10^{-8}$.

@@

The settling times are shown in the lower panel of \autoref{fig:infall}. For recoil velocities $\vk \leq 240\,\kmps$ the kick is not large enough to displace the SMBH beyond the binary-scoured core and the SMBH is considered settled in less than $100\,\mathrm{Myr}$ after SMBH binary coalescence. 
As the kick velocity is increased further, the settling time begins to increase exponentially. 
The slight deviations from an exponential increase are caused by numerical noise, which prevents the median velocity being below $V_\mathrm{RMS}$.

The SMBHs which received a kick velocity larger than $1020\,\kmps$ did not settle back to the centre of the galaxy within the maximum simulation time of $3t_\mathrm{dyn} = 3\,\mathrm{Gyr}$, and already reach maximum displacements in excess of $>60\,\mathrm{kpc}$ during their first excursion from the centre. 
We expect that any SMBH with such a large displacement from its host galaxy would not be observationally identified with that galaxy, and thus we do not include simulations with $\vk>1020\,\kmps$ in this analysis. The used limit for maximum displacement is a few times larger than the largest suggested offset of $19.4\,\mathrm{kpc}$ from observations \citep{barrows2016}.
We do however analyse the $\vk=2000\,\kmps$ simulation, which is above the escape velocity of the galaxy, as a proxy for an SMBH that is almost instantaneously removed from the simulation.
Importantly, an escaping SMBH never completes a single oscillation about the centre of the galaxy.
